# Replication multiregional population projection Willekens & Rogers (1978)

```{r, echo=FALSE, warning=FALSE, message=FALSE}
setwd('/home/mhermans/projects/mstls/')
source('R/mslt.r')
source('R/project.r')
source('R/utils.r')
```


Central steps multiregional/state population projection model:

1. Calculation of observed transfer rates rates $M_x$. 
2. Calculation of $P_x$, the per-region observed death and outmigration probabilities (transfer probabilities), based on $M_x$.
3. Calculation of expected number of survivors, based on $P_x$.
4. Calculation of duration of residence/years lived, $L_x$ based on the number of survivors.
5. Calculation of the surivivorship proportions $S_x$, based on $L_x$.
6. Calculation of the fertility proportions $B_x$, based on $P_x$, $S_x$ and the observed fertility rates.
7. Construction of generalized Leslie matrix $G$ based on $S_x$ and $B_x$.
8. Projection using initial population $n_0$ and $G$.
9. Deriving the stable equivalent population using eigenvalue decomposition of $G$.

## 1. Observed population characteristics

```{r, echo=FALSE}

# read in population Slovenia, table 1.1
# ---------------------------------------
SL <- data.frame(matrix(
  c(
    0, 67800, 0, 417, 0, 192,
    5, 74100, 0, 32, 0, 170,
    10, 70700, 5, 21, 0, 105,
    15, 60100, 953, 31, 0, 310,
    20, 62900, 4444, 47, 0, 451,
    25, 66500, 4204, 45, 0, 368,
    30, 67100, 2758, 67, 0, 252,
    35, 62900, 1438, 77, 0, 111,
    40, 39500, 308, 76, 0, 40,
    45, 47900, 34, 171, 0, 26,
    50, 51300, 15, 268, 0, 34,
    55, 46100, 0, 369, 0, 29,
    60, 39600, 0, 513, 0, 35,
    65, 29500, 0, 763, 0, 28,
    70, 21700, 0, 1036, 0, 19,
    75, 14400, 0, 1088, 0, 16,
    80, 7100, 0, 1041, 0, 5,
    85, 3600, 0, 733, 0, 4), nrow=18, byrow=TRUE))

rownames(SL) <- SL[,1]
SL <- SL[,2:6] # drop ages column
colnames(SL) <- c('population', 'births', 'deaths', 'to_SL', 'to_RYU')

# read in population for rest of Yugoslavia, table 1.1
# -----------------------------------------------------
RYU <- matrix(c(
  0, 847900, 0, 19051, 231, 0,
  5, 905200, 0, 606, 150, 0,
  10, 808100, 54, 386, 127, 0,
  15, 617400, 16335, 534, 419, 0,
  20, 725500, 63828, 885, 680, 0,
  25, 774000, 57477, 1227, 392, 0,
  30, 728400, 32261, 1277, 255, 0,
  35, 633300, 14903, 1313, 143, 0,
  40, 392400, 4729, 1127, 72, 0,
  45, 437100, 940, 1700, 41, 0,
  50, 453800, 324, 2896, 59, 0,
  55, 389300, 0, 3743, 80, 0,
  60, 325800, 0, 5492, 66, 0,
  65, 230600, 0, 6407, 36, 0,
  70, 180000, 0, 8652, 14, 0,
  75, 120900, 0, 8715, 12, 0,
  80, 61200, 0, 6843, 12, 0,
  85, 39300, 0, 5639, 3, 0), nrow=18, byrow=TRUE)

rownames(RYU) <- RYU[,1]
RYU <- RYU[,2:6] # drop ages column
colnames(RYU) <- c('population', 'births', 'deaths', 'to_SL', 'to_RYU')

# joint population for Yugoslavia, cf. table 1.6a
# ------------------------------------------------
YO <- SL + RYU

# merge region migration to arrivals/departures
YO[,4] <- rowSums(YO[,4:5])
YO[,5] <- YO[,4]

colnames(YO) <- c('population', 'births', 'deaths', 'arrivals', 'departures')

```


```{r, echo=FALSE, eval=FALSE}
SL
SL.tot <- colSums(SL)
SL.tot

RYU
RYU.tot <- colSums(YO)
RYU.tot

RYU
colSums(RYU)/colSums(RYU)[1]*5
round((colSums(RYU)/colSums(RYU)[1])*1000,3) # crude observed rates

```

```{r, echo=FALSE}

# Replicate Table 1.3 "Observed rates", pg. 11-12

# observed death rate
DR <- cbind(
  SL[,'deaths']/SL[,'population'],
  RYU[,'deaths']/RYU[,'population'])

# observed fertility rate
FR <- cbind(
  SL[,'births']/SL[,'population'],
  RYU[,'births']/RYU[,'population'])

# observed outmigration rate

MR <- cbind(
  SL[,'to_RYU'] / SL[,'population'],
  RYU[,'to_SL'] / RYU[,'population'])

colnames(MR) <- colnames(FR) <- colnames(DR) <- c('slovenia', 'r.yogos')

# Merge DR, MR into single "long" transition matrix
M.obs <- cbind(
  rbind(
    cbind(rep(0, 18), MR[,1]),
    cbind(MR[,2], rep(0, 18))
    ),
  c(DR[,1], DR[,2])
  )

M.obs <- cbind(
  rep(seq(0,85,5),2),
  rep(seq(4,89,5),2),
  c(rep(1,18), rep(2,18)),
  M.obs)

colnames(M.obs) <- c('x', 'xn', 'i', 'Mx_1i', 'Mx_2i', 'Mxd_i')

```

Replicate Table 1.3 "Observed rates", pg. 11-12

**Observed death rate**

```{r}
round(DR, 6)
```

**Observed fertility rate**

```{r}
round(FR, 6)
```

**Observed migration rate**

```{r}
round(MR, 6)
```



## 2. The multiregional life table

Table 2.1 (p. 22) presents $P_x$, the **probabilities of dying and outmigrating** (or transfer probabilities). The calculation is presented in section 2.7, based on the observed death and migration rates (section 1).

```{r}
M <- transfer_matrix(M.obs, multiple=TRUE)
P <- transfer_prob(M, multiple=TRUE)

ages <- seq(0,85,5)
states <- c('slovenia', 'r.yugos')

P.sl <- state_table(P, 1, ages)
P.ryu <- state_table(P, 2, ages)

# calculate per region death rate
P.sl <- cbind(1 - rowSums(P.sl), P.sl)
P.ryu <- cbind(1 - rowSums(P.ryu), P.ryu)

colnames(P.sl) <- colnames(P.ryu) <- c('death', 'to slov', 'to r.yug')

P.sl # prob. of dying & outmigrating for Slovenia
P.ryu # prob. of dying & outmigrating for rest of Yugoslavia 
```


### 2.1 Life Histories

### 2.2 Expected Number of Survivors at Exact Age x

The expected number of survivors is calculated based on $P_x$, the probability of dying and outmigrating, using

$$
l_{x+5} = P_x l_x
$$

where $l_0$ is a selected cohort size, identical for both regions, in this case 100000.

The following two table replicate tabel 2.3, pg. 30.

```{r}
L.surv <- expected_survivors(P, radix=100000)

# Exp. # of surv., initial region Slov.
round(state_table(L.surv,1, ages, states)) 

# Exp. # of surv., initial region r. Yugos.
round(state_table(L.surv,2, ages, states)) 
```


### 2.3 Duration of Residence and Age Composition of the Life Table Population

The duration of residence by place of birth is calculated for every age interval using

$$
L_x = \frac{5}{2} (\hat{l}_x + \hat{l}_{x+5})
$$

where $l_x$ and $l_{x+5}$ refer to the expected number of survivors for age $x$ (cf. section 2.2). For the last age interval, which is half-open and thus $\hat{l}_{x+5}$ not defined, we use

$$
L_z = M_z^{-1} \hat{l}_z
$$

where $M_z$ is the matrix containg the observed transtion rates for the last age interval.

```{r}

L.dur <- years_lived(L.surv, M)

 # duration/number of years lived in each region, for Slov.
state_table(L.dur,1, ages, states)

# duration/number of years lived in each region, for r. Yugos.
state_table(L.dur,2, ages, states) 

```


### 4. Total number of Years Lived beyond Age x

### 5. Expectation of Life

### 6. Survivorship and Outmigration Proportions

For the projection matrix, we need the age-specific matrices of survivorship proportions $S_x$. These are based on the proportion of survivers $L_x$, using 

$$
S_x = L_{x+5} L_x^{-1}
$$

These age-specific surivivorship proportions for both regions are represented in Table 2.9 (pg. 46-47), replicated below.

```{r}

S <- survivor_prop(L.dur)

# survivorship proportions for Slov.
state_table(S,1, head(ages, -1), states) 

# survivorship proportions for r. Yugos.
state_table(S,2, head(ages, -1), states) 

```


### 2.7 Estimation of Age-specific Outmigration and Death Probabilities

There are tree options for calculating the probabilities of dying and outmigrating (based on the observed transition rates):

* Option 1: No multiple transitions (Rogers, 1975)
* Option 2: Rogers, 1975 (mentioned, not used)
* Option 3: Allows multiple transitions, used.

We start by constructing $M_x$, containing observed outmigration and death rates. Based on this matrix, we can calculated $P_x$ using

$$
P_x = (I + 5/2 M_x)^{-1} (I - 5/2 M_x)
$$

The results are presented in section 2, table 2.1 (supra).


## 3. Multiregional Population Projection

### 3.1 The Growth Matrix

The growth matrix $G$ is a generalized Leslie matrix, constructed from submatrices $S_x$, containing the survivorship proportions (calculated in section 2.6) and $B_s*, containing the fertility rates. E.g, in this two-region example:

$$
M_t = \begin{bmatrix}
0 & 0 & B_{\alpha-5} & B_{\alpha} & B_{\alpha + 5} & B_{\ldots} & B_{\beta - 5} & 0 & 0\\ 
S_{0} & 0 & 0 & 0 & 0 & 0 & 0 & 0 & 0\\ 
0 & S_{5} & 0 & 0 & 0 & 0 & 0 & 0 & 0\\ 
0 & 0 & S_{10} & 0 & 0 & 0 & 0 & 0 & 0\\ 
0 & 0 & 0 & S_{15} & 0 & 0 & 0 & 0 & 0\\ 
0 & 0 & 0 & 0 & S_{20} & 0 & 0 & 0 & 0\\ 
0 & 0 & 0 & 0 & 0 & S_{25} & 0 & 0 & 0\\ 
0 & 0 & 0 & 0 & 0 & 0 & S_{\ldots} & 0 & 0\\ 
0 & 0 & 0 & 0 & 0 & 0 & 0 & S_{z-5} & 0
\end{bmatrix}
$$

with

$$
S_x = \begin{bmatrix}
s^x_{11} & s^x_{21}\\ 
s^x_{12} & s^x_{22}
\end{bmatrix}
$$

$$
B_x = \begin{bmatrix}
b^x_{11} & b^x_{21}\\ 
b^x_{12} & b^x_{22}
\end{bmatrix}
$$

While we have already the $S_x$ submatrices, we still need to calculate $B_x$, based on the observed fertilitary rates $F_x$, the transition probabilites $P_x$ and the survivorship proportions $S_x$, using

$$
B_x = \frac{5}{4} ( P_0 + I ) ( F{x+5} S_x )
$$

```{r}
B <- birth_prop(FR, P, S)

# birth proportions Slov.
round(state_table(B,1, head(ages, -1), states), 6) 

# birth proportions r. Yugo.
round(state_table(B,2, head(ages, -1), states), 6) 

# TODO: (rounding) error last two values RYU?
```

We construct $G$ by combining $B_x$ and $S_x$, resulting in a 36x36 matrix.

```{r}
G <- projection_matrix(B, S)
dim(G)
```



### 3.2 The projection process

The initial population is the observed population at $t_0$, formated as a vector of length 36 with 2 regions nested within 18 age-groups.

```{r}
n0 <- c(SL[,1], RYU[,1])
length(n0)
```

we recursively multiply $G$ and $n0$, projecting the population forwards for 8 steps.

```{r}
result <- project(init_pop=n0, n_states=2, pmat=G, n_steps=8)

proj.slov <- result[1:18,]
proj.ryog <- result[19:36,]
rownames(proj.slov) <- rownames(proj.ryog) <- seq(0,85,5)
colnames(proj.slov) <- colnames(proj.ryog) <- seq(1961,2001,5)

proj.slov # Projection for Solvenia, 1961-2001
proj.ryog # Projection for rest of Yugoslavia, 1961-2001.

```

Replication final subtable table 3.2, pg. 67:

**Projected population distribution 2001**

```{r}
proj.t01 <- cbind(proj.slov[,9], proj.ryog[,9])
proj.t01 <- cbind(rowSums(proj.t01), proj.t01)
colnames(proj.t01) <- c('total', 'slovenia', 'r.yogos')
round(proj.t01)
```

**Projected percentage population distribution 2001**

```{r}
round(prop.table(proj.t01, 2) * 100, 4)
```


### 3.3 The stable equivalent population

We can approximate the stable equivalent to the original population by projecting $n0$ foward a sufficiently large number of steps. This reproduces the percentage distribution of the stable equivalent population, shown in Table 3.3. (p. 70).

```{r}
t500 <- project(n0, n_states=2, pmat=G, n_steps=500)
round(prop.table(matrix(t500[,ncol(t500)], ncol=2), 2)* 100, 4)
```



# Custom functions

## Multistate lifetable functions

```{r}
transfer_matrix
transfer_prob
expected_survivors
years_lived
survivor_prop
birth_prop
```

## Projection functions

```{r}
projection_matrix
project
plot_proj
```


## Helper functions

```{r}
state_table
collapse_interval
```


